Making Higher-Order Superposition Work

### 2022-03-12 ho-techniques/ho-techniques.tex
pre

In 2019 we tested if the idea of gracefully extending first-order provers to
higher-order logic really improves state of the art for the first time. We implemented \lsup{} \cite{bbtvw-21-sup-lam} in
Zipperposition~1.5 with basic heuristics and rudimentary extensions of the
calculus to deal with Booleans. It finished third at that year's higher-order
division of CASC competition \cite{gs-19-casc27}, 12~percentage points behind the
winner, the tableau prover Satallax 3.4 \cite{cb-2013-satallax}.

@@

The main drawback of both $\lambda$-superposition variants compared with
combinatory superposition is that they rely on rules that enumerate possibly
infinite sets of unifiers. We describe a mechanism that interleaves infinitely
branching inferences with the standard saturation process
(Sect.~\ref{sec:ho-tech:infinite-branching}). The prover retains the same
behavior as before on first-order problems, smoothly scaling with increasing
numbers of higher-order clauses.
%
We also propose some heuristics to curb the explosion induced by highly
prolific calculus rules (Sect.~\ref{sec:ho-tech:explosiveness}).

@@

>>>>

In 2019 we tested if the idea of gracefully extending first-order provers to
higher-order logic really improves the state of the art for the first time. We implemented \lsup{} \cite{bbtvw-21-sup-lam} in
Zipperposition~1.5 with basic heuristics and rudimentary extensions of the
calculus to deal with Booleans. It finished third at that year's higher-order
division of CASC competition \cite{gs-19-casc27}, 12~percentage points behind the
winner, the tableau prover Satallax 3.4 \cite{cb-2013-satallax}.

@@

\looseness=-1
The main drawback of both $\lambda$-superposition variants compared with
combinatory superposition is that they rely on rules that enumerate possibly
infinite sets of unifiers. We describe a mechanism that interleaves infinitely
branching inferences with the standard saturation process
(Sect.~\ref{sec:ho-tech:infinite-branching}). The prover retains the same
behavior as before on first-order problems, smoothly scaling with increasing
numbers of higher-order clauses.
%
We also propose heuristics to curb the explosion induced by highly
prolific calculus rules (Sect.~\ref{sec:ho-tech:explosiveness}).

@@

\section{Background and Setting}
\label{sec:ho-tech:background}

We focus on monomorphic higher-order logic, without the axiom of infinity or the
axiom of at least two individuals \cite{bm-14-automation-ho}. However, the
techniques can easily be extended with polymorphism. Indeed, Zipperposition
already supports some of them polymorphically. Further, we use exactly the same
notation for this logic as introduced in Chapter \ref{ch:pre}. Since we are
working with extensions of superposition, we assume the clausal structure
(Sect.~\ref{sec:pre:clauses}). Like in the previous chapter, literals
of clauses can contain arbitrary higher-order terms, including formulas.

\ourpara{Higher-Order Calculi}
\looseness=-1
We briefly introduced the \osup{} calculus \cite{bbtv-21-full-ho-sup} in
Sect.~\ref{sec:pre:ho-sup-calculi}. It is a refutationally complete inference
system and redundancy criterion for higher-order logic with rank-1 polymorphism,
Hilbert choice, and functional and Boolean extensionality.
% The calculus relies on
% \emph{complete sets of unifiers}
% (\emph{CSUs}). The CSU for $s$ and $t$ with respect to a finite set of variables
% $V$, denoted by $\mathrm{CSU}_V(s,t)$, is a set of unifiers of $s$~and~$t$ such
% that for any unifier $\varrho$ of $s$~and~$t$, there exist substitutions $\sigma
% \in \mathrm{CSU}_V(s,t)$ and $\theta$ such that $\varrho(X) = \sigma(\theta(X))$
% for all variables $X \in V$. The set $V$ is used to distinguish
% important variables from auxiliary variables (which may arise in intermediary
% states of the unification procedure). We usually omit it.
Unlike \lsup{}, this calculus 
does not require axioms defining the logical symbols to cope with formulas.
Instead, it includes Boolean inference rules that mimic
superposition from such axioms into Boolean subterms,
while avoiding the explosion incurred by adding these axioms to the proof state. It
also includes rules that simulate Boolean inferences below applied variables.
Both sets of rules are disabled or replaced with incomplete, ad hoc rules
described in the previus chapter in most configurations
of the CASC portfolio.
A new feature of the calculus that we explore in detail is
the ability to select Boolean subterms
to restrict Boolean and superposition inferences.

In contrast to both $\lambda$-superposition variants, combinatory superposition
does not require enumerating elements of CSU to compute results of inferences.
Instead, it avoids computing CSUs by using a form of first-order unification.
Essentially, it enumerates higher-order terms using rules that instantiate
applied variables with partially applied combinators from the complete
combinator set $\{\cst{S}, \cst{K}, \cst{B}, \cst{C}, \cst{I}\}$. This calculus
is the basis of Vampire~4.5 \cite{br-20-full-sup-w-combs}, which finished
closely behind Satallax 3.4 %~and~3.5
at higher-order division of CASC in 2020.

\looseness=-1
A different, very successful calculus is Satallax's SAT-guided tableaux
\cite{backes-brown-2011}. Satallax was the leading higher-order prover of the
2010s. Its simple and elegant tableaux avoid deep superposition-style rewriting
inferences.
Nevertheless, our working hypothesis for the past years has been
that superposition would likely provide a stronger basis for higher-order
reasoning.
Other competing higher-order calculi include SMT (implemented in CVC4
\cite{brotb-19-ho-smt, cbetal-11-cvc4}) and extensional paramodulation (implemented in Leo-III \cite{sb-21-leo3}).

\ourpara{Zipperposition}
Zipperposition \cite{sc-15-simon-phd,bbtvw-21-sup-lam} is a higher-order
theorem prover that implements both $\lambda$-superposition variants, combinatory
superposition, and other superposition-like calculi.
The prover was conceived as a testbed for rapidly
experimenting with extensions of first-order superposition, but over time it
has assimilated many of E's techniques and heuristics and become quite powerful.
Several of our techniques extend the \emph{given clause procedure}, the standard
saturation procedure described in Sect.~\ref{sec:pre:saturation}.

\ourpara{Experimental Setup}
To assess our techniques, we carried out experiments with Zipperposition~2. We
used two sets of benchmarks:\ all 2851~monomorphic higher-order problems from the
TPTP library \cite{gs-17-tptp} version~7.4.0 (labeled \emph{TPTP})
and 1253 Sledgehammer-generated
monomorphic higher-order problems (labeled \emph{SH}).
Although some techniques support polymorphism, we
uniformly used monomorphic benchmarks.

We fixed a \emph{base} configuration
of Zipperposition parameters as a baseline for all comparisons. This is an
incomplete, pragmatic configuration of \osup{} using heuristics expected to perform
well on a wide range of problems.
%Note that
%we use a different baseline configuration than in our earlier paper \cite{making-ho-work}.
%%% That goes without saying. We use a different calculus! Now clarified in intro. --JB
In each experiment, we varied
the parameters associated with a specific technique to evaluate it. The
experiments were run on StarExec Miami \cite{sst-14-starexec} servers, equipped with
Intel Xeon E5-2620 v4 CPUs clocked at 2.10 GHz. Unless otherwise stated, we used a
CPU time limit of 15~s, roughly the time each configuration is given in the
CASC portfolio mode. The raw evaluation results are available online.%
\footnote{\url{http://doi.org/10.5281/zenodo.5007440}}

### 2022-03-12 scratch/paper.tex
pre

\newcommand\NumberOK[1]{#1}
\newcommand\NumberNOK[1]{\colorbox{red}{#1}}
\newcommand{\ourparagraph}[1]{\paragraph{\upshape\bfseries#1}}

@@

\ourparagraph{Higher-Order Logic}
We define terms $s, t, u, v$ inductively as free variables $F, X$, bound
variables $x, y, z, \dotsc$, constants $\cst{f}, \cst{g},\allowbreak
\cst{a},\allowbreak \cst{b}, \dotsc$, term applications $s \, t$, and
$\lambda$-abstractions $\lamx{s}$. The syntactic distinction between free and
bound variables yields \emph{loose bound variables} (e.g., $y$ in $\lamx{y \,
\cst{a}}$) \cite{tn-93-patterns}.
%We shorten iterated application $s \, t_1 \, \cdots \, t_n$ to $s \,
%\overline{t}_n$ and $\lambda$-abstraction $\lambda x_1. \, \cdots \, \lambda
%x_n. \> s$ to $\lam{\overline{x}_n}{s}$.
We let $s \, \overline{t}_n$ stand for $s \, t_1 \, \ldots \, t_n$ and
$\lam{\overline{x}_n}{s}$ for $\lambda x_1. \ldots \lambda x_n. \> s$. The $n$-fold application of
a unary term $s$ to a term $t$ is denoted by $s^n \, t$. Every
$\beta$-normal term can be written as $\lam{\overline{x}_m}{s \,
\overline{t}_n}$, where $s$ is not an application; we call $s$ the \emph{head}
of the term. If the type of a term $t$ is of the form $\tau_1 \to \cdots \to
\tau_n \to o$, where $o$ is the distinguished Boolean type and $n \ge 0$, we
call $t$ a \emph{predicate}. A term of type $o$ is called a \emph{formula}.

@@

\ourparagraph{Higher-Order Calculi}
\looseness=-1
The \osup{} calculus \cite{bbtv-21-full-ho-sup} is a refutationally
complete inference system and redundancy criterion for higher-order logic with
rank-1 polymorphism, Hilbert choice, and functional and Boolean extensionality.
The calculus relies on
\emph{complete sets of unifiers}
(\emph{CSUs}). The CSU for $s$ and $t$ with respect to a finite set of variables
$V$, denoted by $\mathrm{CSU}_V(s,t)$, is a set of unifiers of $s$~and~$t$ such
that for any unifier $\varrho$ of $s$~and~$t$, there exist substitutions $\sigma
\in \mathrm{CSU}_V(s,t)$ and $\theta$ such that $\varrho(X) = \sigma(\theta(X))$
for all variables $X \in V$. The set $V$ is used to distinguish
important variables from auxiliary variables (which may arise in intermediary
states of the unification procedure). We usually omit it.

@@

\ourparagraph{Zipperposition}
Zipperposition \cite{sc-15-simon-phd,bbtvw-21-sup-lam} is a higher-order
theorem prover that implements both \lsup{} variants, combinatory
superposition, and other superposition-like calculi.
The prover was conceived as a testbed for rapidly
experimenting with extensions of first-order superposition, but over time it
has assimilated many of E's techniques and heuristics and become quite powerful.

@@

\ourparagraph{Experimental Setup}
To assess our techniques, we carried out experiments with Zipperposition~2. We
used two sets of benchmarks:\ all 2851~monomorphic higher-order problems from the
TPTP library \cite{gs-17-tptp} version~7.4.0 (labeled \emph{TPTP})
and 1253 Sledgehammer-generated
monomorphic higher-order problems (labeled \emph{SH}).
Although some techniques support polymorphism, we
uniformly used monomorphic benchmarks.

@@

\ourparagraph{Evaluation and Discussion}

@@

\ourparagraph{Evaluation and Discussion}
  %
  % \begin{enumerate}
  %   \item Different points when E is called, for different time periods
  %   \item Number of clauses from $M_\mathrm{ho}$ that is translated
  %   \item Different kinds of liftings
  % \end{enumerate}

>>>>

\newcommand\NumberOK[1]{#1}
\newcommand\NumberNOK[1]{\colorbox{red}{#1}}
\newcommand{\ourpara}[1]{\paragraph{\upshape\bfseries#1}}

@@

\ourpara{Higher-Order Logic}
We define terms $s, t, u, v$ inductively as free variables $F, X$, bound
variables $x, y, z, \dotsc$, constants $\cst{f}, \cst{g},\allowbreak
\cst{a},\allowbreak \cst{b}, \dotsc$, term applications $s \, t$, and
$\lambda$-abstractions $\lamx{s}$. The syntactic distinction between free and
bound variables yields \emph{loose bound variables} (e.g., $y$ in $\lamx{y \,
\cst{a}}$) \cite{tn-93-patterns}.
%We shorten iterated application $s \, t_1 \, \cdots \, t_n$ to $s \,
%\overline{t}_n$ and $\lambda$-abstraction $\lambda x_1. \, \cdots \, \lambda
%x_n. \> s$ to $\lam{\overline{x}_n}{s}$.
We let $s \, \overline{t}_n$ stand for $s \, t_1 \, \ldots \, t_n$ and
$\lam{\overline{x}_n}{s}$ for $\lambda x_1. \ldots \lambda x_n. \> s$. The $n$-fold application of
a unary term $s$ to a term $t$ is denoted by $s^n \, t$. Every
$\beta$-normal term can be written as $\lam{\overline{x}_m}{s \,
\overline{t}_n}$, where $s$ is not an application; we call $s$ the \emph{head}
of the term. If the type of a term $t$ is of the form $\tau_1 \to \cdots \to
\tau_n \to o$, where $o$ is the distinguished Boolean type and $n \ge 0$, we
call $t$ a \emph{predicate}. A term of type $o$ is called a \emph{formula}.

@@

\ourpara{Higher-Order Calculi}
\looseness=-1
The \osup{} calculus \cite{bbtv-21-full-ho-sup} is a refutationally
complete inference system and redundancy criterion for higher-order logic with
rank-1 polymorphism, Hilbert choice, and functional and Boolean extensionality.
The calculus relies on
\emph{complete sets of unifiers}
(\emph{CSUs}). The CSU for $s$ and $t$ with respect to a finite set of variables
$V$, denoted by $\mathrm{CSU}_V(s,t)$, is a set of unifiers of $s$~and~$t$ such
that for any unifier $\varrho$ of $s$~and~$t$, there exist substitutions $\sigma
\in \mathrm{CSU}_V(s,t)$ and $\theta$ such that $\varrho(X) = \sigma(\theta(X))$
for all variables $X \in V$. The set $V$ is used to distinguish
important variables from auxiliary variables (which may arise in intermediary
states of the unification procedure). We usually omit it.

@@

\ourpara{Zipperposition}
Zipperposition \cite{sc-15-simon-phd,bbtvw-21-sup-lam} is a higher-order
theorem prover that implements both \lsup{} variants, combinatory
superposition, and other superposition-like calculi.
The prover was conceived as a testbed for rapidly
experimenting with extensions of first-order superposition, but over time it
has assimilated many of E's techniques and heuristics and become quite powerful.

@@

\ourpara{Experimental Setup}
To assess our techniques, we carried out experiments with Zipperposition~2. We
used two sets of benchmarks:\ all 2851~monomorphic higher-order problems from the
TPTP library \cite{gs-17-tptp} version~7.4.0 (labeled \emph{TPTP})
and 1253 Sledgehammer-generated
monomorphic higher-order problems (labeled \emph{SH}).
Although some techniques support polymorphism, we
uniformly used monomorphic benchmarks.

@@

\ourpara{Evaluation and Discussion}

@@

\ourpara{Evaluation and Discussion}
  %
  % \begin{enumerate}
  %   \item Different points when E is called, for different time periods
  %   \item Number of clauses from $M_\mathrm{ho}$ that is translated
  %   \item Different kinds of liftings
  % \end{enumerate}
